Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-11-10 ms/review-responses.tex

\reply{
We have added a figure (Figure \ref{fig:runtimes}) and short discussion \llname{runtimesxn} of run times.
\plr{what's up with those run times?}
}

\begin{point}{\revref}
    I read the acknowledgement to the Hearth and Creative Sky Brewing with a sense of familiarity in feeling of gratitude to my own favorite cafes and breweries, but I it's not a great precedent for Acknowledgements to be filled this way. Please cut.
\end{point}

\reply{
    Good point; we have done this.
}

\begin{point}{Figure \ref{fig:ibs}:}
    Show random-mating expectation
\end{point}

\reply{
    This is done.
}

\begin{point}{Figure \ref{fig:sumstats}A, \ref{fig:allsfs}:}
    Perhaps more revealing to show on log-log scale?
\end{point}

\reply{
    Good suggestion -- the SFS in Fig \ref{fig:sumstats}A is now on a log-log scale,
    which shows the slight decrease in low frequency SNPs a little better. 
    We've left Fig \ref{fig:allsfs} as-is.
}

@@

\reply{
This is a good point also raised by other reviewers. We have added an appendix section comparing our model to a reverse-time stepping stone simulation on several relevant summary statistics \llname{appendix1}. We show that for $\theta_\pi$ and several other statistics the stepping stone model approaches the continuous model as the resolution of the landscape increases. We also see some interesting differences that seem to reflect artifacts from discretization. We are also curious how all the other analyses we test would be affected using other simulation schemes, but hope to explore that aspect in future work. 
}

\begin{point}{}
    Related to the first point, to what extent are the results qualitatively different to those that would be obtained in a stepping-stone model? My interpretation is that they are actually very similar, but I didn't see whether that was explicitly discussed. In some sense, it's still easier to do large simulations in a stepping-stone model so it would be nice to be reassured that that's still ok.
\end{point}

\reply{
    See above. It is certainly easier and faster, but comes with significant issues either when neighborhood size is lower than the population of a deme, or when demes are small enough that the sample size approaches local $N_e$, at least in reverse-time simulations. 
}

\begin{point}{}
    The source of equation \eqref{eqn:p_survival} is not obvious to me. I sort of see how it makes sense, but a little but more intuition or a brief derivation or an illuminating either in the main text or the supplement, would be helpful.
\end{point}

### 2019-11-10 ms/spaceness.tex
a bit more

To avoid questionable assumptions, we here used forward-time, individual-based simulations.
The question of what regulates real populations has a long history and many answers
\citep{lloyd1967crowding,antonovics1980ecological},
but it is clear that populations must have some density-dependent feedback on population size,
or else they would face eventual extinction or explosion.
In the absence of unrealistic global population regulation,
this regulation must be local,
and there are many ways to arrange this \citep{bolker2003spatial}.
We decrease each individual's probability of survival with increasing local population density,
which shifts reproductive output towards low-density regions 
(and prevents the unrealistic population clustering seen by \citet{Felsenstein1975}). 
Such models have been used extensively in ecological modeling
\citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression,}
but rarely in population genetics, \revpoint{AE}{2}
where to our knowledge implementations of continuous space models
before their availability through SLiM \citep{Haller2019}
have focused on a small number of genetic loci
\citep{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation},
which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

To maintain a stable population, mortality increases with local population density.
To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$,
where $g$ is the Gaussian density with mean zero and standard deviation $\sigma$.
Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
where $d_{ij}$ is the distance between individuals $i$ and $j$ and the sum is over all neighbors within distance $3 \sigma$.
Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
Then, given a per-unit carrying capacity $K$, \llabel{ll:K_defn}
the probability of survival until the next time step for individual $i$ is
\begin{equation}
    p_i 
    =
    \min\left( 0.95, \frac{1}{1 + n_i / (K (1+L))} \right) .
\end{equation}
We chose this functional form so that the equilibrium population density per unit area is close to $K$, \llabel{ll:K_again}
and the mean lifetime is around $L$.

@@

Adjusting the spatial dispersal and interaction distance, $\sigma$,
has a surprisingly large effect on demographic quantities that are usually fixed in Wright-Fisher models --
the generation time, census population size, and variance in offspring number, shown in Figure~\ref{fig:genparams}. Because our simulation is paramaterized on an individual level, these population parameters emerge as a property of the interactions among individuals rather than being directly set \llabel{emergentparams}.
Variation across runs occurs because,
even though the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more}
were the same in all simulations,
the strength of stochastic effects depends strongly on the spatial interaction distance $\sigma$.
For instance, 
the population density near to individual $i$ (denoted $n_i$ above)
is computed by averaging over roughly $N_W = 4 \pi K \sigma^2$ individuals,
and so has standard deviation proportional to $1/\sqrt{N_W}$ --
it is more variable at lower densities.
(Recall that $N_W$ is Wright's neighborhood size.)
Since the probability of survival is a nonlinear function of $n_i$,
actual equilibrium densities and lifetimes differ from $K$ and $L$.
This is the reason that we included \emph{random mating} simulations --
where mate choice and offspring dispersal are both nonspatial --
since this should preserve the random fluctuations in local population density
while destroying any spatial genetic structure.
We verified that random mating models retained no geographic signal
by showing that summary statistics did not differ significantly between sampling regimes
(Table~\ref{table:sampling}), unlike in spatial models (discussed below).

@@

Spatial structure is particularly challenging for genome-wide association studies, because the effects of dispersal on genetic variation are compounded by spatial variation in the environment \citep{Mathieson2012}. Spatially restricted mate choice and dispersal causes variation in allele frequencies across the range of a species. If environmental factors affecting the phenotype of interest also vary over space, then allele frequencies and environmental exposures will covary over space. In this scenario an uncorrected GWAS will infer genetic associations with a purely environmental phenotype at any site in the genome that is differentiated over space, and the relative degree of bias will be a function of the degree of covariation in allele frequencies and the environment (i.e., Figure \ref{fig:gwas}C, bottom panel). This pattern has been demonstrated in a variety of simulation and empirical contexts \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015,Berg2018,Sohail2018}.

@@

\bibliography{spaceness}
\bibliography{more_refs}

@@

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/wm_v_dd_maps.pdf}
\caption{Maps of individual locations in a continuous-space Wright-Malécot model with independent dispersal of all individuals (top) and under our continuous space model incorporating density-dependant fitness (bottom). The clustering seen in the top row is the "Pain in the Torus" described by \citet{Felsenstein1975}.}
\label{fig:slim_v_msp_sfs} 
\end{figure*}
\FloatBarrier

\begin{figure*}[p]
\centering
\includegraphics[]{figures/slim_v_msp_sfs.pdf}
\caption{Site frequency spectra from a simulation with neighborhood size = 12.5 when mutations are recorded directly in SLiM (blue line) or applied later in msprime (red line).}
\label{fig:slim_v_msp_sfs} 
\end{figure*}
\FloatBarrier

>>>>

To avoid hard-to-evaluate approximations, we here used forward-time, individual-based simulations.
The question of what regulates real populations has a long history and many answers
\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population},
but it is clear that populations must at some point have density-dependent feedback on population size,
or else they would face eventual extinction or explosion.
In the absence of unrealistic global population regulation,
this regulation must be local,
and there are many ways to arrange this \citep{bolker2003spatial}.
In our simulations,
we decrease each individual's probability of survival with increasing local population density,
which shifts reproductive output towards low-density regions 
(and prevents the unrealistic population clustering seen by \citet{Felsenstein1975},
see Supplemental \autoref{fig:slim_has_no_pain}).
Such models have been used extensively in ecological modeling
\citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression}
but rarely in population genetics, \revpoint{AE}{2}
where to our knowledge implementations of continuous space models
before their availability through SLiM \citep{Haller2019}
have focused on a small number of genetic loci
\citep[e.g.,][]{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation,rossine2014espao},
which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

To maintain a stable population, mortality increases with local population density.
To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$,
where $g$ is the Gaussian density with mean zero and standard deviation $\sigma$.
Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
where $d_{ij}$ is the distance between individuals $i$ and $j$ and the sum is over all neighbors within distance $3 \sigma$.
Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
Then, given a per-unit carrying capacity $K$, \llabel{ll:K_defn}
the probability of survival until the next time step for individual $i$ is
\begin{equation} \label{eqn:p_survival}
    p_i 
    =
    \min\left( 0.95, \frac{1}{1 + n_i / (K (1+L))} \right) .
\end{equation}
We chose this functional form so that the equilibrium population density per unit area is close to $K$, \llabel{ll:K_again}
and the mean lifetime is around $L$;
for more description see Appendix \ref{apx:demographic_model}.

@@

Adjusting the spatial dispersal and interaction distance, $\sigma$,
has a surprisingly large effect on demographic quantities that are usually fixed in Wright-Fisher models --
the generation time, census population size, and variance in offspring number, shown in Figure~\ref{fig:genparams}. Because our simulation is paramaterized on an individual level, these population parameters emerge as a property of the interactions among individuals rather than being directly set \llabel{emergentparams}.
Variation across runs occurs because,
even though the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more} \revpoint{1}{11}
were the same in all simulations,
the strength of stochastic effects depends strongly on the spatial interaction distance $\sigma$.
For instance, 
the population density near to individual $i$ (denoted $n_i$ above)
is computed by averaging over roughly $N_W = 4 \pi K \sigma^2$ individuals,
and so has standard deviation proportional to $1/\sqrt{N_W}$ --
it is more variable at lower densities.
(Recall that $N_W$ is Wright's neighborhood size.)
Since the probability of survival is a nonlinear function of $n_i$,
actual equilibrium densities and lifetimes differ from $K$ and $L$.
This is the reason that we included \emph{random mating} simulations --
where mate choice and offspring dispersal are both nonspatial --
since this should preserve the random fluctuations in local population density
while destroying any spatial genetic structure.
We verified that random mating models retained no geographic signal
by showing that summary statistics did not differ significantly between sampling regimes
(Table~\ref{table:sampling}), unlike in spatial models (discussed below).

@@

Spatial structure is particularly challenging for genome-wide association studies, because the effects of dispersal on genetic variation are compounded by spatial variation in the environment \citep{Mathieson2012}. Spatially restricted mate choice and dispersal causes variation in allele frequencies across the range of a species. If environmental factors affecting the phenotype of interest also vary over space, then allele frequencies and environmental exposures will covary over space. \llabel{ll:1:11:2}
In this scenario an uncorrected GWAS will infer genetic associations with a purely environmental phenotype at any site in the genome that is differentiated over space, and the relative degree of bias will be a function of the degree of covariation in allele frequencies and the environment (i.e., Figure \ref{fig:gwas}C, bottom panel). This pattern has been demonstrated in a variety of simulation and empirical contexts \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015,Berg2018,Sohail2018}.

@@

\bibliography{spaceness,more_refs}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Demographic model}
\label{apx:demographic_model}

We chose our demographic model so that every individual has on average $1/L$ offspring
each time step,
and if the local population density of an individual is $n$,
then their probability of survival until the next time step is
(equation \eqref{eqn:p_survival}):
\begin{equation}
    p = \min\left( 0.95, \frac{1}{1 + n / (K (1+L))} \right) .
\end{equation}
We capped survival at 0.95 so that we would not have exceptionally long-lived individuals
in sparsely populated areas --
otherwise, an isolated individual might live for a very long time.
Since $1 - p \approx n / (K (1 + L))$,
mortalitity goes up roughly linearly with the number of neighbors (on a scale given by $K$),
as would be obtained if, for instance, mortality is due to agonistic interactions.
Ignoring in/outmigration,
a region is at demographic equilibrium if the per-capita probability of death
is equal to the birth rate,
i.e., if $1 - p = 1/L$.
(Note that there is no effect of age in the model,
which would make the analysis more complicated.)
Solving this for $n$,
we get that in a well-mixed population, the equilibrium density should be around
\begin{equation}
    n &= K \frac{L+1}{L-1}
\end{equation}
individuals per unit area.
At this density, the per-capita death rate is $1/L$, so the mean lifetime is $L$.
This equilibrium density is \emph{not} $K$, but (since $L=4$) is two-thirds larger.
However, in practice this model leads to a total population size which is around $K$
multiplied by total geographic area
(but which depends on $\sigma$, as discussed above).
The main reason for this is that
since offspring tend to be near their parents, individuals tend to be ``clumped'',
and so experience a higher average density than the ``density'' one would compute
by dividing census size by geographic area \citep{lloyd1967crowding}.
To maintain a constant expected total population size
would require making (say) $K$ depend on $\sigma$;
however, typical local population densities might then be more dissimilar.

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/wm_v_dd_maps.pdf}
\caption{Maps of individual locations in a continuous-space Wright-Malécot model with independent dispersal of all individuals (top) and under our continuous space model incorporating density-dependant fitness (bottom). The clustering seen in the top row is the "Pain in the Torus" described by \citet{Felsenstein1975}.}
\label{fig:slim_has_no_pain} 
\end{figure*}
\FloatBarrier

\begin{figure*}[p]
\centering
\includegraphics[]{figures/slim_v_msp_sfs.pdf}
\caption{Site frequency spectra from a simulation with neighborhood size = 12.5 when mutations are recorded directly in SLiM (blue line) or applied later in msprime (red line).}
\label{fig:slim_v_msp_sfs2} 
\end{figure*}
\FloatBarrier
